---
title: "Open Access Evidence in Unpaywall"
description: |
  We investigated more than 31 million scholarly journal articles from 2008 - 2018 contained in Unpaywall, a widely used open access discovery tool. Using Google BigQuery and R, we determined around 11 million articles (37%), where Unpaywall links to open access full-texts. Our data analysis revealed various open access location and evidence types, as well as large overlaps between them, raising important questions about how to responsibly re-use Unpaywall data in bibliometric research and open access monitoring.
author:
  - name: Najko Jahn 
    url: https://twitter.com/najkoja
    affiliation: State and University Library Göttingen
    affiliation_url: https://www.sub.uni-goettingen.de/
  - name: Anne Hobert
    affiliation: State and University Library Göttingen
    affiliation_url: https://www.sub.uni-goettingen.de/
date: "`r Sys.Date()`"
creative_commons: CC BY
output: distill::distill_article
bibliography: pubs.bib
draft: true
---

```{r, echo = FALSE, message = FALSE, warning = FALSE}
knitr::opts_chunk$set(
  comment = "#>",
  collapse = TRUE,
  warning = FALSE,
  message = FALSE,
  echo = TRUE
)
options(scipen = 999, digits = 2)
knitr::knit_hooks$set(
  inline = function(x) {
    if (is.numeric(x)) {
      return(prettyNum(x, big.mark = ","))
    } else{
      return(x)
    }
  }
)
```

[Unpaywall](https://unpaywall.org/), developed and maintained by the [team of Impactstory](http://impactstory.org/team), finds open access copies of scholarly literature [@Piwowar_2018]. Providing DOIs, [Unpaywall's REST API](https://unpaywall.org/products/api) not only returns open access full-text links, but also helpful metadata about the open access status of publications indexed in [Crossref](https://www.crossref.org/). While the API allows to retrieve a limited amount of records, Unpaywall also offers [database snapshots](https://unpaywall.org/products/snapshot) for large-scale analysis, which more and more bibliometric databases and open access monitoring services utilise. However, documentations about how they work with these dumps are hard to find.

In this blog post, we show how we made use of the Unpaywall data dump in [Google BigQuery](https://cloud.google.com/bigquery/), a cloud-based service that allows fast analysis of large datasets, and how we interfaced BigQuery for our analysis with R. We wanted to know the extent of open access status information in Unpaywall, particularly, how this information can be utilised for bibliometric research. In our case, we intend to match evidence from Unpaywall with the Web of Science in-house database from the [German Competence Center for Bibliometrics](http://www.bibliometrie.info/) to determine factors influencing open access publication activities among German Universities as part of our [BMBF-funded research project OAUNI](https://www.wihoforschung.de/de/oauni-2182.php).

## Store and analyse large datasets with Google BigQuery 

## Unpaywall Overview

In R, we interface our Unpaywall dataset stored in Google BigQuery with the packages [DBI](https://www.r-dbi.org/) and [bigrquery](https://github.com/r-dbi/bigrquery).


```{r fetch_bq}
#' connect to google bg where we imported the jsonl Unpaywall dump
library(DBI)
library(bigrquery)
con <- dbConnect(
  bigrquery::bigquery(),
  project = "api-project-764811344545",
  dataset = "oadoi_full"
)
```

Our BigQuery project has two tables, one containing all records between 2008 - 2012, and another for more recent works being published since 2013. When connecting with `tbl()`, Google asks us to login via a web browser or to supply an private access token to interface our access-protected database. 

```{r}
library(tidyverse)
upw_08_12 <- tbl(con, "feb_19_mongo_export_2008_2012_full_all_genres")
upw_13_19 <- tbl(con, "feb_19_mongo_export_2013_Feb2019_full_all_genres")
```

[bigrquery](https://github.com/r-dbi/bigrquery) allows querying BigQuery tables using SQL or [dplyr](https://dplyr.tidyverse.org/) functions. The latter is convenient when you just started to learn SQL, but feel more experienced in the [tidyverse](https://www.tidyverse.org/), a popular collection of R packages following the Wickham-Grolemund approach to practise data science [@Wickham_2017]. Here's an example where we call BigQuery with dplyr, which is part of the tidyverse, to obtain the first ten records from 2018. We restrict our search to journal articles, the most common genre in Unpaywall.

```{r}
upw_13_19 %>%
  filter(year == 2018, genre == "journal-article") %>%
  head(10)
```

Notice that [our schema](bq_schema.json) follows the [Unpaywall data format](https://unpaywall.org/data-format). However, we excluded the large data fields `z-authors` from our dataset. Moreover, we did not consider the fields `title`, `doi-url`, which is redundant to the `doi` field, and `best-oa-location`, which is derived from the Open Access location object.  

In this blog post, we will examine the following open access indicators: 

- `is_oa`: A logical value indicating whether an open accessible version of the article was found or not.

- `journal_is_in_doaj`: A logical value indicating whether an article was published in a journal registered in the [Directory of Open Access Journals (DOAJ)](https://doaj.org/).

The column  `oa_locations` is a [list-column](https://jennybc.github.io/purrr-tutorial/ls13_list-columns.html) that contains individual metadata about all open access full-text links found per article. By definition, open access provision is not limited to one route, but multiple copies of an article can be made freely available at the same time using various means [@Suber_2012].

Here are the three data variables from the `oa_locations` object that we will analyse as well:

- `is_best`: A logical value defined by Unpaywall's algorithm that describes the most relevant open access location. The algorithm prioritizes publisher-hosted content.

- `host_type`: Is the open access full-text provided by a publisher or a repository? 

- `evidence`: How the open access full-text was found by Unpaywall?

### Open Access availability (`is_oa`)

To start with, we retrieve the number and proportion of journal articles with open access full-text published between 2008 and 2018 using Unpaywall's most basic open access indicator `is_oa`, a logical value, which is `TRUE` when at least one open access full-text was found. After matching and summarizing the `is_oa` observations by year with dplyr, `collect()` loads the so aggregated data from BigQuery into a local tibble.

```{r cache=TRUE}
oa_08_12 <- upw_08_12 %>%
  # query and aggregate with dpylr 
  filter(genre == "journal-article") %>%
  group_by(year, is_oa) %>%
  summarise(n = n()) %>% 
  # load the data into a local tibble
  collect()
oa_13_18 <- upw_13_19 %>%
  # query and aggregate with dpylr
  filter(genre == "journal-article", year < 2019) %>%
  group_by(year, is_oa) %>%
  summarise(n = n()) %>% 
  # load the data into a local tibble
  collect()
my_df <- bind_rows(oa_08_12, oa_13_18) %>%
  # calculate proportion per year
  ungroup() %>%
  mutate(year = lubridate::ymd(paste0(year, "-01-01"))) %>%
  group_by(year, is_oa) %>%
  summarise(n = sum(n)) %>%
  mutate(prop = n / sum(n))
my_df
```

In total, `r sum(my_df$n)` journal articles published between 2008 - 2018 were included in Unpaywall. For `r  my_df %>% filter(is_oa == TRUE) %>% .$n %>% sum()` articles, Unpaywall was able to link a DOI to at least one freely available full-text (`r  my_df %>% filter(is_oa == TRUE) %>% .$n %>% sum() / sum(my_df$n) * 100` %). In conclusion, around every third scholarly journal article published since 2008 is currently openly available. 

Next, let's plot the prevalence of open access to journal articles over year using the data visualisation package [ggplot](https://ggplot2.tidyverse.org/index.html). To make our ggplot object interactive, we turn it into a [plotly](https://plot.ly/) chart, a javascript library, using `ggplotly()`. The tooltip presents the total number and percentage for each category and year.

```{r, layout="l-body-outset", fig.cap="Open Access to Journal Articles according to Unpaywall. Blue area represents journal articles with at least one freely available full-text, the gray area represents toll-access articles."}
library(scales)
plot_a <- my_df %>%
  # prepare label that we want to present as tooltip
  mutate(`Proportion in %` = round(prop * 100, 2)) %>%
  ggplot(aes(year, n, label = `Proportion in %`)) +
  geom_area(aes(fill = is_oa, group = is_oa),  alpha = 0.8) +
  labs(x = "Year published", y = "Journal Articles",
       title = "Open Access to Journal Articles") +
  scale_fill_manual("Is OA?",
                    values = c("#b3b3b3a0", "#56B4E9")) +
  scale_x_date(date_labels = "%y") +
  scale_y_continuous(labels = scales::number_format(big.mark = " ")) +
  theme_minimal(base_family = "Roboto") +
  theme(plot.margin = margin(30, 30, 30, 30)) +
  theme(panel.grid.minor = element_blank()) +
  theme(axis.ticks = element_blank()) +
  theme(panel.grid.major.x = element_blank()) +
  theme(panel.border = element_blank())
# turn ggplot object into interactive plotly chart
plotly::ggplotly(plot_a, tooltip = c("label", "y")) 
```


While a general growth of journal articles and open access provision to them can be observed, there is a considerable decline in the number of journal articles published in 2018, presumably because of an indexing lag between Crossref and Unpaywall. The decline in open access full-text availability was even clearer, suggesting that some open access content is provided after a certain period of time. 


### Unpaywall Open Access Hosting Types (`host_type`)

Using Unpaywall's open access location types allows for a more detailed analysis of open access provision. In the following, we explore the variable `host_type`, showing whether Unpaywall found the open access full-text on a publisher website or in a repository. Furthermore, we include articles from fully open access journals that are indexed in [Directory of Open Access Journals (DOAJ)](https://doaj.org/) as indicated by the `journal_is_in_doaj` variable in our analysis. As a start, we only examine the best open access location per DOI, `is_best`. As said before, this variable is defined by Unpaywall's algorithm that prioritizes publisher-hosted content.

Instead of dplyr, we are now querying BigQuery with SQL. Before, we built and tested the SQL queries in the BigQuery web UI. The SQL code is stored in separate files ([host_type_08_12.sql](host_type_08_12.sql) and [host_type_13_18.sql](host_type_13_18.sql)), which are shared in the source code repository of this blog:

```{r}
host_type_08_12_query <- readLines("host_type_08_12.sql") %>%
  paste(collapse = "")
host_type_13_18_query <- readLines("host_type_13_18.sql") %>%
  paste(collapse = "")
```

After calling BigQuery using SQL with the [DBI](https://www.r-dbi.org/) interface, we bind the two resulting data frames into one. Using `case_when()`from dpylr, we create a `host` column distinguishing between "DOAJ-listed Journal", "Other Journals" and "Repositories only" open access provision.

```{r}
host_type_08_12_query_df <- dbGetQuery(con, host_type_08_12_query)
host_type_13_18_query_df <- dbGetQuery(con, host_type_13_18_query)
host_type_df <-
  bind_rows(host_type_08_12_query_df, host_type_13_18_query_df) %>%
  mutate(
    host = case_when(
      journal_is_in_doaj == TRUE ~ "DOAJ-listed Journal",
      host_type == "publisher" ~ "Other Journals",
      host_type == "repository" ~ "Repositories only"
    )
  ) %>%
  mutate(year = lubridate::ymd(paste0(year, "-01-01")))
host_type_df
```

To explore our data, we follow [Claus Wilke's excellent book "Fundamentals of Data Visualization"](https://serialmentor.com/dataviz/)[@Wilke_2019] and [visualise our proportions separately as parts of the total](https://serialmentor.com/dataviz/visualizing-proportions.html#visualizing-proportions-separately-as-parts-of-the-total). Again, our final ggplot graphic is transformed to an interactive plotly chart.

```{r, layout="l-body-outset", fig.cap='Open Access to journal articles by  open access hosting location. The colored bars represent the number of open access articles per host ("DOAJ-listed Journal", "Other Journals", "Repositories only"), the gray bar the total number of journal articles indexed in Crossref where Unpaywall was able to identify at least one open access full-text.'}
# calculate all oa articles per year
all_articles <- host_type_df %>%
  ungroup() %>%
  group_by(year) %>%
  summarise(number_of_articles = sum(number_of_articles))

plot_b <-
  ggplot(host_type_df, aes(x = year, y = number_of_articles)) +
  geom_bar(
    data = all_articles,
    aes(fill = "All OA Articles"),
    color = "transparent",
    stat = "identity"
  ) +
  geom_bar(aes(fill = "by Host"), color = "transparent", stat = "identity") +
  facet_wrap( ~ host, nrow = 1) +
  scale_fill_manual(values = c("#b3b3b3a0", "#56B4E9"), name = "") +
  labs(x = "Year", y = "OA Articles (Total)") +
  theme(legend.position = "top",
        legend.justification = "right") +
  scale_x_date(date_labels = "%y") +
  scale_y_continuous(labels = scales::number_format(big.mark = " ")) +
  theme_minimal(base_family = "Roboto") +
  theme(plot.margin = margin(30, 30, 30, 30)) +
  theme(panel.grid.minor = element_blank()) +
  theme(axis.ticks = element_blank()) +
  theme(panel.grid.major.x = element_blank()) +
  theme(panel.border = element_blank())
# turn ggplot object into interactive plotly chart
plotly::ggplotly(plot_b, tooltip = c("y")) 
```

The figure shows that most publisher-provided open access links were obtained from journals that were not indexed in the DOAJ (`r host_type_df %>% filter(host == "Other Journals") %>% .$number_of_articles %>% sum()` articles, representing `r host_type_df %>% filter(host == "Other Journals") %>% .$number_of_articles %>% sum() / host_type_df %>% .$number_of_articles %>% sum() * 100` % of all journal articles with openly available full-text identified by Unpaywall). 

While the number publications in DOAJ-indexed journals is rising constantly, open access provided by other journal types and repositories declined from 2017 to 2018. Indeed, there is a considerable amount of journals that delay open access provision [@Laakso_2013]. A prominent example is the journal [Cell](https://www.cell.com/cell/archive) where all articles are made freely available after an embargo period of twelve months. Also self-archiving in repositories is often subject to embargo periods imposed by publishers, or is done after publication [@Bj_rk_2013]. A more detailed analysis of delayed open access, however, is challenging using Unpaywall data only, because it is not tracked when articles were made open access.

### Unpaywall Open Access Evidence Types (`evidence`)

Evidence can be used to further detail open access provision by host_type.

## Overlap of Open Access Provision and Evidence Types

Unpaywall not only describes one, but all open access full-texts it discovers with useful metadata. In the following, we will analyse if and to which extent the various open access indicators intersect. We start with an analysis of the overlap between host tpyes, followed by determining set intersections between Unpaywall's evidence types.

### Overlap between Host Types

So far, we only examined the best open access location per DOI, indicated by `is_best`, a variable defined by Unpaywall algorithm that prioritizes publisher-hosted content. To present articles that are both provided by publishers and repositories, we use the BigQuery SQL function [STRING_AGG](https://cloud.google.com/bigquery/docs/reference/standard-sql/functions-and-operators#string_agg) to make a new variable where we concatenate the different host_types per open access article (for more details, see the full SQL queries [`host_type_intersect_08_12.sql`](host_type_intersect_08_12.sql) and [`host_type_intersect_13_18.sql`](host_type_intersect_13_18.sql)].

```{r}
host_type_intersect_08_12_query <- readLines("host_type_intersect_08_12.sql") %>%
  paste(collapse = "")
host_type_intersect_13_18_query <- readLines("host_type_intersect_13_18.sql") %>%
  paste(collapse = "")
```

Again, we call BigQuery and load the so aggregated data into our local R session for further data exploration. This time, we do not want to present the total number of open access publications, but is relative share. We already obtained the total number of articles stored in `my_df`.

```{r}
host_type_08_12_intersect_df <-
  dbGetQuery(con, host_type_intersect_08_12_query)
host_type_13_18_intersect_df <-
  dbGetQuery(con, host_type_intersect_13_18_query)
host_type_intersect <-
  bind_rows(host_type_08_12_intersect_df, host_type_13_18_intersect_df) %>%
  mutate(year = lubridate::ymd(paste0(year, "-01-01"))) %>%
  mutate(
    host = case_when(
      host_type_count == "publisher" ~ "Publisher only",
      host_type_count == "publisher,repository" ~ "Publisher & Repository",
      host_type_count == "repository" ~ "Repositories only"
    )
  ) %>%
  mutate(host = factor(
    host,
    levels = c("Publisher only", "Publisher & Repository", "Repositories only")
  ))
#' obtain yearly publication volumes
host_type_intersect <- my_df %>%
  group_by(year) %>%
  summarise(all_articles = sum(n)) %>%
  # join with host type figures
  right_join(host_type_intersect, by = "year") %>%
  # calculate proportion
  mutate(prop = number_of_articles / all_articles)
host_type_intersect
```

Let's visualise host type distribution including the overlap between publisher and repository provided open access:

```{r, layout="l-body-outset", fig.cap='Open Access to journal articles by open access hosting location. The colored bars represent the number of open access articles per Unpaywall host category: "publisher" and "repository", the gray bar the percentage of open access to journal articles indexed in Crossref from Unpaywall. Because open access provision is not mutually exclusive, the overlap between "publisher" and "repository" hosted open access full-texts is also shown'}
# get overall oa share
host_type_all <- host_type_intersect %>%
  group_by(year) %>%
  summarise(prop = sum(prop))
# make a ggplot graphic
plot_host_intersect <-
  ggplot(host_type_intersect, aes(x = year, y = prop)) +
  geom_bar(
    data = host_type_all,
    aes(fill = "All OA Articles"),
    color = "transparent",
    stat = "identity"
  ) +
  geom_bar(aes(fill = "by Host"), color = "transparent", stat = "identity") +
  facet_wrap(~ host, nrow = 1) +
  scale_fill_manual(values = c("#b3b3b3a0", "#56B4E9"), name = "") +
  labs(x = "Year", y = "OA Share",
       title = "Overlap between Open Access Host Types in Unpaywall") +
  theme(legend.position = "top",
        legend.justification = "right") +
  scale_x_date(date_labels = "%y") +
  scale_y_continuous(labels = scales::percent_format(accuracy = 5L)) +
  theme_minimal(base_family = "Roboto") +
  theme(plot.margin = margin(30, 30, 30, 30)) +
  theme(panel.grid.minor = element_blank()) +
  theme(axis.ticks = element_blank()) +
  theme(panel.grid.major.x = element_blank()) +
  theme(panel.border = element_blank())
# turn ggplot object into interactive plotly chart
plotly::ggplotly(plot_host_intersect, tooltip = c("y")) 
```

The Figure shows that Unpaywall found most open access full-text links on publisher websites, of which a large proportion was not archived on a repository. The overlap of open access provided by both routes deserves attention: a proportion of around `r filter(host_type_intersect, host != "Publisher only") %>% .$number_of_articles %>% sum() / host_type_intersect %>% .$number_of_articles %>% sum() * 100` % of all open access articles was provided in open access using both publisher web sites and repositories.

### Overlaps between Evidence types


## Discussion and Conclusion

In this blog post, we demonstrated how to analyse the Unpaywall data dump with Google BigQuery and R. Interfacing BigQuery with R has allowed us to integrate a high-performance and user-friendly database environment into our R data analytics workflow. Using this data management environment, we found 11 million journal articles published between 2008-18 with open access full-texts, representing around 1/3 of all articles indexed in Crossref for this period. Moreover, we found Unpaywall being a suitable data source for open access analytics, because it does not only tag if a publication is freely available, but also contains metadata describing how and where the various open access full-text links were discovered. 

Our Unpaywall data analysis revealed various open access location and evidence types, as well as large overlaps between them. Along with the likely influence of embargoed or delayed open access provision on some of these types, our analysis raises important questions about how to responsibly use Unpaywall data in bibliometric research and open access monitoring. Examining Unpaywall's best open access location only, favors publisher-provided open access, which, in turn, means that open access provided by repositories like the [arXiv](https://arxiv.org/) would be underestimated. Likewise, large overlaps between evidence categories can be observed. To allow for careful consideration, bibliometric research and open access monitoring must therefore be clear about how open access indicators were derived from Unpaywall.

In future, we will use these insights from our data analysis to work on a matching procedure between Unpaywall and the Web of Science in-house database from the [German Competence Center for Bibliometrics](http://www.bibliometrie.info/) in our [OAUNI project](https://www.wihoforschung.de/de/oauni-2182.php). In doing so, we want to represent Unpaywall's open access evidence as comprehensive as possible to allow for a more pluralist view on open access to journal articles from researchers affiliated with German Universities.

## Acknowledgments {.appendix}

We acknowledge financial support from the the Federal Ministry of Education and Research of Germany (BMBF) in the framework [Quantitative research on the science sector](https://www.wihoforschung.de/en/quantitative-research-on-the-science-sector-1573.php) (Project: "OAUNI Entwicklung und Einflussfaktoren des Open-Access-Publizierens an Universitäten in Deutschland – Teilvorhaben Erhebung und Evaluierung", Förderkennzeichen: 01PU17023A)

