---
title: "How open are hybrid journals included in nationwide transformative agreements in Germany?"
description: "We present hoaddata, an experimental R package that combines open scholarly data from Crossref and OpenAlex. Using this package, we illustrate the progress made in publishing open access content across hybrid journal portfolios and countries."
author:
  - name: Najko Jahn 
    url: https://twitter.com/najkoja
    affiliation: State and University Library Göttingen
    affiliation_url: https://www.sub.uni-goettingen.de/
date: May 27, 2022
output: distill::distill_article
always_allow_html: true
bibliography: literature.bib
---

```{r, echo = FALSE, message = FALSE, warning = FALSE}
knitr::opts_chunk$set(
  comment = "#>",
  collapse = TRUE,
  warning = FALSE,
  message = FALSE,
  echo = FALSE
)
options(scipen = 999, digits = 2)
knitr::knit_hooks$set(
  inline = function(x) {
    if (is.numeric(x)) {
      return(prettyNum(x, big.mark = ","))
    } else{
      return(x)
    }
  }
)
```

Assessing the volume and share of open access articles in hybrid journals is crucial for the planning and implementation of [transformative agreements](https://esac-initiative.org/about/transformative-agreements/), an evolving type of  contracts between libraries and publishers where subscription spending is repurposed for open access publishing [@Schimmer_2015]. 
However, library consortia that mainly negotiate such agreements with large commercial publishers often lack these data.
Here, we present [hoaddata](https://subugoe.github.io/hoaddata/), an experimental R package, in which openly available journal-level data about nationwide transformative agreements in Germany are combined with article-level open access status information and country affiliations. 
Accordingly, hoadata provides essential data for the monitoring and benchmarking of transformative agreements across hybrid journal portfolios and countries. <!--, aiming at better understanding the current state of the transition of subscription-based publishing to fully open access.-->

Interacting with data through R packages like hoaddata makes data analytics more transparent because R packages meet generic principles for computational reproducibility: coherent file organisation, separation of data, methods and results, and specification of the computational environment [@Marwick_2018].
For data science practitioners, R packages, thus, provide a reliable way to re-use data and code.
In our specific case, hoaddata not only contains datasets about hybrid open access. It also comprises code used to compile the data by interfacing our cloud-based [Google Big Query data warehouse](https://subugoe.github.io/scholcomm_analytics/data.html), where we store open scholarly data from [Crossref](https://www.crossref.org/), [OpenAlex](https://openalex.org/) and [Unpaywall](https://unpaywall.org/).
hoaddata is automatically built and updated with GitHub Actions.
Each merge event into the main branch triggers the execution of code, ensuring that the most recent snapshots available in our data warehouse will be used to obtain data about nationwide transformative agreements.
Data changes including updates will be incorporated in the package and tracked with Git, which makes it possible to reproduce different version of the data contained in hoaddata. 

In this blog post, we describe the data analytics workflow behind hoaddata. 
The main purpose of hoaddata is to ship data for monitoring dashboards about the progress of nationwide transformative agreements, which we currently develop in the [HOAD project with the support of the Deutsche Forschungsgemeinschaft](https://gepris.dfg.de/gepris/projekt/416115939), but everyone can install the package from GitHub and use it in R.
To demonstrate its potential, we will use hoaddata to illustrate the current state of the transition of hybrid journals to fully open access relative to those journals, which are included in nationwide transformative agreements in Germany. 


## Data and methods

hoaddata focuses on nationwide transformative agreements in Germany. As a first step, we draw on the work of the [German Open Access Monitor (OAM)](https://open-access-monitor.de/) to obtain a list of journals under these agreements [@VTQXLM_2021].
We merged all journals into a single data file and enriched it with missing ISSN variants. 
Because publishers register journal-level metadata in Crossref when they first deposit metadata for a given journal including ISSNs, we furthermore matched the OAM journal list with [Crossref's title list](https://www.crossref.org/documentation/reports/browsable-title-list/) to link ISSN variants to journals as they were represented in Crossref metadata.

<aside>This [thread](https://community.crossref.org/t/parallel-titles-for-a-given-issn/2183) is helpful to better understand journal curation workflows in Crossref.</aside>

After obtaining a list of hybrid journals linked to nationwide transformative agreements, we determined the article volume by journal and year using Crossref. Following Unpaywall's approach, Crossref metadata records considered as front matter were excluded.^[<https://support.unpaywall.org/support/solutions/articles/44001894783-what-does-is-paratext-mean-in-the-api->] 
Open access articles were identified through Creative Commons license URLs in Crossref metadata records.
License URL were mapped to the different license versions like CC BY. 

Because country affiliations are a key data point for nationwide transformative agreements, we used OpenAlex to determine the country share per journal and publisher portfolio.
To our knowledge, OpenAlex does not provide information about corresponding authors and their affiliation.
Instead, we made use of first-author affiliations.
A first author is often regarded as being the [lead author](https://en.wikipedia.org/wiki/Lead_author) who has usually undertaken most of the research presented in the article, although author roles can vary across disciplines.
In case OpenAlex did not record any country affiliation, we extracted country names from the metadata field `display_name` using regular expressions.
We applied full counting to account for multiple country affiliations.

As a result, hoaddata  provides the following datasets:

- [oam_hybrid_jns](https://subugoe.github.io/hoaddata/reference/oam_hybrid_jns.html): Hybrid journals included in the Open Access Monitor. Data were gathered from @VTQXLM_2021, validated and mapped to Crossref-indexed journals.

- [cc_jn_ind](https://subugoe.github.io/hoaddata/reference/cc_jn_ind.html): Prevalence of Creative Commons license variants by year and hybrid journal as obtained from Crossref.

- [cc_openalex_inst_jn_ind](https://subugoe.github.io/hoaddata/reference/cc_openalex_inst_jn_ind.html): First author country affiliations per journal, year and Creative Commons license. Country affiliations were gathered from OpenAlex.

Article-level data

- [cc_openalex_inst](https://subugoe.github.io/hoaddata/reference/cc_openalex_inst.html): Article-level affiliation data from first authors as obtained from OpenAlex. Covers only open access articles under a Creative Commons license in a hybrid journal. 

hoaddata can be installed from [GitHub](https://github.com/subugoe/hoaddata):

``` r
# install.packages("remotes")
remotes::install_github("subugoe/hoaddash", dependencies = "Imports")
```

You can also directly download the data as csv files from [GitHub](https://github.com/subugoe/hoaddata/).
The files stored in the [`data-raw`](https://github.com/subugoe/hoaddata/tree/main/data-raw) folder of the package together with the code used to create the datasets. Specific SQL queries can be found in [`inst/sql/`](https://github.com/subugoe/hoaddata/tree/main/inst/sql).


## Use-Case: Open access uptake in hybrid journals included in nationwide transformative agreements in Germany

```{r, echo=FALSE}
library(hoaddata)
library(dplyr)
```

At the time of writing, hoaddata comprised information about `r length(unique(hoaddata::cc_jn_ind$cr_journal_id))` hybrid journals included in twenty consortial transformative agreements in Germany. 
Since 2017, these journals have published `r length(unique(hoaddata::cc_openalex_inst$doi))` open access articles with an Creative Commons License, representing an share of `r round(length(unique(hoaddata::cc_openalex_inst$doi)) / hoaddata::cc_openalex_inst_jn_ind %>% distinct(cr_journal_id, cr_year, articles_total) %>% pull(articles_total) %>% sum(), 3) * 100`%. 

### Open access uptake in Germany's nationwide transformative agreements in 2021

Using OpenAlex's affiliation information, we can break down the performance of transformative agreements to countries. 
Showing the publication year 2021, the following interactive table compares the global publication volume including open access with that of German lead authors by hybrid journal portfolio. 
It highlights the dominant position of transformative agreements negotiated by the [DEAL consortium](https://deal-operations.de/) in terms of articles published and open access.
It is important to note that [DEAL has reached no agreement with Elsevier](https://www.projekt-deal.de/elsevier-news/) so far, one of the largest scholarly publishers, which is, in turn, not included in the analysis.


```{r, message=FALSE, echo=FALSE}

my_jn_df <- oam_hybrid_jns |>
  distinct(agreement, lead, cr_journal_id) |>
  inner_join(cc_jn_ind, by = "cr_journal_id")
## Portflio-specific statistics per year
pubs_per_year <- my_jn_df |>
  distinct(agreement, lead, cr_journal_id, jn_all, cr_year) |>
  group_by(agreement, lead, cr_year) |>
  summarize(n_journals = n_distinct(cr_journal_id),
            articles = sum(jn_all))

cc_pubs_per_year <- my_jn_df |>
  group_by(agreement, lead, cr_year) |>
  summarise(cc_articles = sum(cc_total))

agreement_per_year <- left_join(pubs_per_year, cc_pubs_per_year, 
                                by = c("agreement", "lead", "cr_year")) %>%
  mutate(cc_share = cc_articles / articles)
  
## German lead author statistics per year
de_pubs <- cc_openalex_inst_jn_ind |>
  filter(country_code == "DE") |>
  inner_join(oam_hybrid_jns, by = "cr_journal_id")
de_total <- de_pubs |>
  distinct(agreement, lead, cr_year, cr_journal_id, articles_total) %>%
  group_by(agreement, lead, cr_year) %>%
  summarize(de_n_journals = n_distinct(cr_journal_id),
            de_articles = sum(articles_total))
de_cc <- de_pubs |>
  filter(!is.na(cc)) |>
  distinct(agreement, lead, cr_year, cr_journal_id, articles_under_cc_variant) |>
  group_by(agreement, lead, cr_year) |>
  summarize(cc_de_n_journals = n_distinct(cr_journal_id),
            cc_de_articles = sum(articles_under_cc_variant))

de_df <- left_join(de_total, de_cc, by = c("agreement", "lead", "cr_year")) |>
  mutate(cr_year = factor(cr_year)) %>%
  mutate(de_cc_share = cc_de_articles / de_articles) %>%
  mutate_if(is.numeric, ~replace(., is.na(.), 0))

agreement_tbl_df <- inner_join(agreement_per_year, de_df, 
                               by = c("agreement", "lead", "cr_year"))
```

```{r, echo = FALSE}
#' React bar chart helper
#'
#' From <https://glin.github.io/reactable/articles/building-twitter-followers.html>
#'
#' @importFrom htmltools div
#'
#' @noRd
react_bar_chart <-
  function(label,
           width = "100%",
           height = "14px",
           fill = "#00bfc4",
           background = NULL) {
    bar <-
      htmltools::div(style = list(
        background = fill,
        width = width,
        height = height
      ))
    chart <-
      htmltools::div(style = list(
        flexGrow = 1,
        marginLeft = "6px",
        background = background
      ),
      bar)
    htmltools::div(style = list(display = "flex", alignItems = "center", marginLeft = "10px"), label, chart)
  }
```

```{r, message=FALSE, echo = FALSE}
#' Sparkline widget
#' 
cc_trend <- agreement_tbl_df %>% 
  group_by(agreement, lead) %>%
  summarise(cc_trend = list(round(cc_share * 100, 1)),
            cc_de_trend = list(round(de_cc_share * 100, 1))) %>%
  mutate(cc_trend_spark = NA, cc_de_trend_spark = NA)

agreement_tbl_df <- left_join(agreement_tbl_df, cc_trend, by = c("agreement", "lead")) %>%
  relocate(cc_trend_spark, .after = cc_share) %>%
  relocate(cc_de_trend_spark, .after = de_cc_share) %>%
  mutate(cr_year = as.numeric(as.character(cr_year)))


```

```{r, echo=FALSE}
library(crosstalk)
library(htmltools)

# Initialize shared Crosstalk data

shared_df <- agreement_tbl_df %>%
  # only 2021
  filter(cr_year == "2021") %>% 
  SharedData$new()
```

```{r, echo = FALSE}
library(reactable)
library(sparkline)

g <- reactable::reactable(
  shared_df,
  pagination = TRUE,
  highlight = TRUE,
  # new in v0.2.3.9000
  defaultColDef = colDef(vAlign = "center", headerClass = "header2"),
  defaultSorted = "articles",
  defaultSortOrder = "desc",
  defaultSelected = 1:20,
  style = list(#fontFamily = "Karla",
               whiteSpace = "pre"),
  columns = list(
    agreement = colDef(
      "Agreement",
      cell = function(value, index) {
        lead <- shared_df$data()$lead[index]
        htmltools::tagList(htmltools::div(style = list(
          fontWeight = 600, color = "#333"
        ), value),
        htmltools::div(style = list(fontSize = 10), lead))
      },
      width = 150,
      align = "left",
      style = list(whiteSpace = "pre"),
      sticky = "left"
    ),
    n_journals = colDef(
      "Journals",
      format = colFormat(separators = TRUE, locales = "en-GB"),
      width = 95,
      style = list(whiteSpace = "pre"),
      show = FALSE
    ),
    # Articles
    articles  = colDef(
      "Articles",
      format = colFormat(separators = TRUE, locales = "en-GB"),
      width = 110,
      style = list(whiteSpace = "pre"),
      class = "number border-left",
    ),
    de_articles = colDef(
      name = "Articles",
      format = colFormat(separators = TRUE, locales = "en-GB"),
      width = 110,
      style = list(whiteSpace = "pre"),
      class = "number border-left"
    ),
    # CC Shares Global / Germany
    cc_share = colDef(
      name = "% OA CC license",
      # Render the bar charts using a custom cell render function
      cell = function(value) {
        value <- paste0(format(round(value * 100, 1), nsmall = 1), "%")
        value <- format(value, width = 5, justify = "right")
        react_bar_chart(value,
                        width = value,
                        fill = "#0093c7ff",
                        background = "#e1e1e1")
      },
      align = "center",
      style = list(whiteSpace = "pre"),
      class = "number",
      html = TRUE,
      minWidth = 130
    ),
    de_cc_share = colDef(
      "% OA CC license",
      # Render the bar charts using a custom cell render function
      cell = function(value) {
        value <- paste0(format(round(value * 100, 1), nsmall = 1), "%")
        value <- format(value, width = 5, justify = "right")
        react_bar_chart(value,
                        width = value,
                        fill = "#00c7ab",
                        background = "#e1e1e1")
      },
      align = "center",
      style = list(whiteSpace = "pre"),
      class = "number",
      minWidth = 130
    ),
    # Render trend as sparklines
    cc_trend_spark = colDef(
      "Trend 2017-22",
#      header = with_tooltip("Trend 2017-22", "Percentage of articles under Creative Commons license over time"),
      cell = function(value, index) {
        sparkline(
          shared_df$data()$cc_trend[[index]],
          lineColor = "#0093c7ff",
          fillColor = "transparent",
          highlightSpotColor = "#0093c7ff",
          minSpotColor = FALSE,
          maxSpotColor = FALSE,
          spotColor = FALSE,
          lineWidth = 2.5,
          spotRadius = 4,
          width = 100
        )
      },
      align = "center",
      sortable = FALSE,
      minWidth = 115
    ),
    cc_de_trend_spark = colDef(
      "Trend 2017-22",
      cell = function(value, index) {
        sparkline(
          shared_df$data()$cc_de_trend[[index]],
          lineColor = "#00c7ab",
          fillColor = "transparent",
          highlightSpotColor = "#00c7ab",
                    highlightSpotColor = "#00c7ab",

          minSpotColor = FALSE,
          maxSpotColor = FALSE,
          spotColor = FALSE,
          lineWidth = 2.5,
          spotRadius = 4,
          width = 100
        )
      },
      align = "center",
      sortable = FALSE,
      minWidth = 115
    ),
    # Don't show columns that are used for filtering or calculations
    lead = colDef(show = FALSE),
    cr_year = colDef(show = FALSE),
    de_n_journals = colDef(show = FALSE),
    cc_de_n_journals = colDef(show = FALSE),
    cc_articles = colDef(show = FALSE),
    cc_de_articles = colDef(show = FALSE),
    cc_trend = colDef(show = FALSE),
    cc_de_trend = colDef(show = FALSE)
  ),
  # Create column groups
  columnGroups = list(
    colGroup(
      name = "Global Publication Volume",
      columns = c("articles", "cc_share", "cc_trend_spark"),
      headerClass = "group-header"
    ),
    colGroup(
      name = "German Lead Authors",
      columns = c("de_articles", "de_cc_share", "cc_de_trend_spark"),
      headerClass = "group-header"
    )
  ),
  compact = TRUE,
  defaultPageSize = 6,
  language = reactableLang(
      pageInfo = "{rowStart}\u2013{rowEnd} of {rows} consortial agreements in Germany",
      pagePrevious = "\u276e",
      pageNext = "\u276f",
    ),
elementId = "years-filter-table"
)
```


::: l-body-outset

```{r,echo=FALSE, eval = TRUE}
htmltools::div(
  class = "agreement-tbl",
  # filters,
  htmltools::div(
    class = "filters",
     htmltools::div(
       class = "filter-input",
       filter_select(
         "filter_vertrag",
         "Consortium Lead",
         shared_df,
         ~ lead,
          multiple = TRUE
       )
     ),
  ),
  # table
  g,
  # footer
  htmltools::div(
    class = "agreement-footer",  paste(
      "Data sources: Open Access Monitor Zeitschriftenlisten (v2), Crossref, OpenAlex. Last updated:", Sys.Date(), ".")
  )
)
```

:::

```{r, echo =FALSE}
tags$link(href = "https://fonts.googleapis.com/css?family=Karla:400,700|Fira+Mono&display=fallback", rel = "stylesheet")
```

```{css, echo=FALSE}
.filters {
  display: flex;
  flex-wrap: wrap;
  margin-top: 4px;
  margin-bottom: 8px;
  margin-left: -32px;
}

.filter-input {
  margin-top: 4px;
  margin-left: 32px;
  flex: 1;
  max-width: 250px;
}

.filter-input label {
  color: hsl(0, 0%, 45%);
  font-weight: 700;
  font-size: 13px;
  letter-spacing: 0.4px;
  text-transform: uppercase;
}

.group-header {
  font-weight: 700;
  font-size: 14px;
  letter-spacing: 0.4px;
  text-transform: uppercase;
}

.filter-input select,
.filter-input input[type="search"] {
  padding: 0 6px;
  height: 32px;
}

.filter-input input[type="search"] {
  /* Reset Bootstrap 3 styles */
  -webkit-appearance: searchfield;
}

.filter-input input[type="search"]::-webkit-search-cancel-button {
  /* Reset Bootstrap 3 styles */
  -webkit-appearance: searchfield-cancel-button;
}


.followers-tbl a {
  color: inherit;
}
.header2 {
  border-bottom: 2px solid #555;
  font-size: 14px;
  font-weight: bold;
  text-transform: uppercase;
}
.header:hover {
  background-color: #eee;
}
/* Highlight headers when sorting */
.header:hover,
.header[aria-sort="ascending"],
.header[aria-sort="descending"] {
  background-color: rgba(236, 236, 237, 1);
}
.agreement-footer {
  margin: 18px 0;
  font-size: 10px;
  font-family: "Karla", Helvetica Neue, Helvetica, Arial, sans-serif;
}
.agreement-tbl {
  margin: 18px 0;
  font-size: 14px;
}

.bar-cell {
  display: flex;
  align-items: center;
}

.number {
  font-family: "Fira Mono", Consolas, Monaco, monospace;
  font-size: 13.5px;
  white-space: pre;
}

.bar-chart {
  flex-grow: 1;
  margin-left: 6px;
  height: 14px;
}

.bar {
  height: 100%;
}


.border-left {
  border-left: 1px solid #b5b5b5;
}

.name {
  font-weight: 900;
}
```

```{r}
# Global
all_21 <- cc_jn_ind %>%
  distinct(cr_journal_id, cr_year, jn_all) %>%
  filter(cr_year == 2021) %>%
  pull(jn_all) %>%
  sum()

oa_21 <- cc_jn_ind %>%
  filter(!is.na(cc), cr_year == 2021) %>%
  pull(cc_total) %>%
  sum()

# Germany
all_21_de <- cc_openalex_inst_jn_ind %>%
  filter(country_code == "DE", cr_year == 2021) %>%
  distinct(cr_journal_id, articles_total) %>%
  pull(articles_total) %>%
  sum()

oa_21_de <- cc_openalex_inst_jn_ind %>%
  filter(country_code == "DE", cr_year == 2021, !is.na(cc)) %>%
  pull(articles_under_cc_variant) %>%
  sum()
```

Interestingly, Germany's open access share in 2021 performed in most cases below 80%, suggesting that not all authors made use of open access options, or were not eligible to publish open access, likely because their institution was not part of a consortium.
But also some article types might no be included in agreements.
For instance, many article types in medical specialist journals from Springer Medizin, which are targeted at medical practitioners, are not covered by DEAL.^[See contract <https://pure.mpg.de/rest/items/item_3174351_1/component/file_3189424/content>]
We were not able to obtain Creative Commons license information from Crossref metadata records representing hybrid journal portfolios from ACM, AIP, Hogrefe and SPIE.

The table furthermore shows large discrepancies between the global open access uptake in hybrid journals and Germany's.
In 2021, `r oa_21 / all_21 * 100`% of articles published in hybrid journals included in nationwide transformative agreements in Germany were open access, while the open access percentage among articles published by lead authors based in Germany was `r oa_21_de / all_21_de *100`%. 
Overall, `r oa_21_de / oa_21 * 100`% of open access articles published in hybrid journals included in transformative agreements in 2021 were from Germany, although its total publication volume accounted for `r all_21_de / all_21 * 100`% of articles published.

### Country overview


As can be seen from the following figure, which highlights the Top 20 most productive countries, open access uptake in hybrid journals included in nationwide transformative agreements in Germany varies across countries.
Notably, lead authors based in the US, China or India -- which together accounted for roughly 40% of articles published -- did utilize open access options to a much lesser extent than authors from European countries. 
Together with Germany, the UK, the Netherlands, Sweden and Switzerland have gained an open access share above 50% over the years, most likely because of providing nationwide transformative agreements with similar journal coverage. 


```{r, warning=FALSE, message = FALSE, echo = FALSE}
library(forcats)
library(ggplot2)
library(ggiraph)

country_by_year <- cc_openalex_inst_jn_ind |>
  distinct(cr_year, country_code, cr_journal_id, articles_total) |>
  group_by(cr_year, country_code) |>
  summarise(all = sum(articles_total))

cc_country_by_year <- cc_openalex_inst_jn_ind |>
  filter(!is.na(cc)) |>
  group_by(cr_year, country_code) |>
  summarise(cc = sum(articles_under_cc_variant))

# prepare label and facets
all_pubs <- cc_jn_ind %>% distinct(cr_journal_id, cr_year, jn_all) %>% .$jn_all %>% sum()

# top 20 countries in terms of total publication volume
top_20 <- cc_openalex_inst_jn_ind |>
  distinct(cr_year, country_code, cr_journal_id, articles_total) |>
  filter(!is.na(country_code)) |>
  group_by(country_code) |>
  summarise(n = sum(articles_total)) %>%
  arrange(desc(n)) %>%
  mutate(position = row_number()) %>%
  mutate(
    country_name = countrycode::countrycode(country_code, origin = "iso2c", destination = "country.name")
  ) %>%
  mutate(country_name = ifelse(position <= 20, country_name, "Others")) %>%
  group_by(country_name) %>%
  summarise(n = sum(n)) %>%
  mutate(pub_share = n / all_pubs) %>%
  mutate(country_label = paste0(toupper(country_name), "\n",
                                paste0(format(round(n / 1e3, 0), trim = TRUE), 
                                "k / ", round(pub_share * 100, 1), "%"))) %>%
  arrange(desc(n)) %>%
  mutate(country_name = fct_inorder(country_name)) %>%
  mutate(country_name = fct_relevel(country_name, "Others", after = Inf)) %>%
  arrange(country_name) %>%
  mutate(country_label = fct_inorder(country_label))
  
# relative
country_df_rel <- inner_join(country_by_year, cc_country_by_year, 
                             by = c("cr_year", "country_code")) %>%
  mutate(
    country_name = countrycode::countrycode(country_code, origin = "iso2c", destination = "country.name")
  ) %>%
  mutate(country_name = ifelse(country_name %in% top_20$country_name, country_name, "Others")) %>%
  inner_join(top_20, by = "country_name") %>%
  group_by(cr_year, country_name, country_label) %>%
  summarise(all = sum(all),
           cc = (sum(cc))) %>%
  mutate(prop = cc / all) %>%
  # Highlight Germany
  mutate(my_cols = ifelse(country_name == "Germany", "#00c7ab", "#B0B0B0D0")) %>%
  mutate(cr_year = gsub("^20", "'", cr_year))

# tooltips
# ranges for retangles plot, which we want to use for data hover per period
my_rects <- tibble::tibble(
  starts = as.character(2017:2021),
  ends = as.character(2018:2022),
  group = seq_along(2017:2021),
  my_tooltip = as.character(2018:2022)) |>
  mutate(across(, function(x) gsub("^20", "'", x))) %>%
  inner_join(country_df_rel, by = c("my_tooltip" = "cr_year")) %>%
  mutate(tooltip_text = glue::glue('<small>{country_name}</small><br><b>{format(cc, big.mark = ",")} / {round(prop * 100, 1)}%</b><br><small>articles with CC license in {my_tooltip}</smalL>'))


country_multiples <- ggplot(country_df_rel, aes(cr_year, prop, fill = my_cols, group = my_cols)) +
    geom_area(stat = "identity") +
    geom_rect_interactive(
    data = my_rects,
    inherit.aes = FALSE,
    aes(
      xmin = starts,
      xmax = ends,
      ymin = 0,
      ymax = 0.75,
      group = group,
      tooltip = tooltip_text
    ),
    color = "transparent",
    fill = "transparent",
    alpha = 0.01
  ) +
  facet_wrap(~country_label, nrow = 3) +
  scale_y_continuous(labels = scales::percent_format(accuracy = 1L),
                     expand = expansion(mult = c(0, 0.05)),
                     breaks = c(0, .25, .5, .75)) +
  scale_fill_identity() +
  theme_minimal() +
  labs(y = "OA Percentage", x = "2017 - 2022", 
         title = "Who published open access in hybrid journals\nincluded in nationwide transformative agreements in Germany?\n",
         subtitle = "Country\n Total Lead Author Articles / Global Market Share") +
  theme(panel.grid.major.x = element_blank(),
        panel.grid.minor.x = element_blank(),
        panel.grid.minor.y = element_blank(),
        panel.ontop = FALSE,
        strip.text = element_text(size = rel(.56)),
        axis.text.y = element_text(size = rel(.8)),
        plot.title = element_text(size = rel(1.2)),
        plot.title.position = "plot",
        plot.subtitle = element_text(size = rel(.68), hjust = 0.5),
        panel.spacing = unit(1, "lines"),
        axis.text.x=element_blank())
```

::: l-body-outset

```{r, fig.cap="Percentage of articles with a Creative Commons license per country. Showing the Top 20 most productive countries in terms of articles published by lead authors between 2017 and 2022 in hybrid journals included in nationwide transformative agreements in Germany. Country-specific charts are sorted by the total number of lead author articles.\nData sources: Open Access Monitor Zeitschriftenlisten (v2), Crossref, OpenAlex.", echo=FALSE}
ggiraph::girafe(
  ggobj = country_multiples,
  width_svg = 7,
  height_svg = 7 * 0.618,
 options = list(opts_tooltip(
    css="background-color:white;
;font-size:1.15em;padding:10px;border-radius:5px;box-shadow: 0 0 10px 0 rgba(0, 0, 0, 0.5)",
    opacity = .95)))
```

:::

## Discussion and outlook

Over the past years, Germany's library consortia sucessfully negotiated transformative agreements with commercial publishers, resulting in an increase of open access articles from lead authors based in Germany that were published under an Creative Commons license. 
But this growth of open access is neither balanced across journal portfolios nor across countries. 

As illustrated, there are substantial variations across journal portfolios in terms of open access articles published by lead authors based in Germany. They can be explained by different agreement terms such as the number and types of institutions involved or the restriction of open access publishing options to specific article types. 
In its recent [Reference Guide to Transformative Agreements](https://esac-initiative.org/about/transformative-agreements/reference-guide/implementation/), and in line with previous research [@Borrego_2020], the [ESAC initiative](https://esac-initiative.org/) points out that  agreement terms affecting the scope of contracts can have a large impact on the performance of transformative agreements.

In terms of country variations, although the [ESAC Registry of Transformative Agreements](https://esac-initiative.org/about/transformative-agreements/agreement-registry/) discloses an increasing number of similar national-level agreements, more than 80% of articles published in hybrid journals included in nationwide transformative agreements in Germany are still behind a paywall. 
Tracking open access across country affiliations reveals that uptake rates are particular low in the most productive countries USA, China and India. 
It remains to be seen whether a transition of hybrid journal portfolios to fully open access through transformative agreements is feasible given these observed global differences.

In future, we want to use hoaddata as data source for monitoring dashboards about the progress of nationwide transformative agreements in Germany, which we currently develop in the HOAD project with the support of the Deutsche Forschungsgemeinschaft.
We want to extend its current scope on data about publication volume and open access uptake across agreements and countries. 
The aim is to highlight open metadata gaps not only relative to Creative Commons license information. 
Following up on our work on [metacheck](https://subugoe.github.io/metacheck/), an email tool to check metadata compliance, we will also present information about the coverage of Text and Data Mining support, funder infos, ORCIDs or Open Abstracts in Crossref metadata records; these data are critical for an open and reproducible monitoring of transformative agreements.

The launch of dashboards specific to nationwide transformative agreements in Germany is targeted for September 2022.



