---
title: "Analysing and re-classifying open access information in OpenAlex"
output: distill::distill_article
date: "2023-11-07"
author:
  - name: Najko Jahn 
    affiliation: State and University Library Göttingen
    affiliation_url: https://www.sub.uni-goettingen.de/
    orcid_id: 0000-0001-5105-1463
  - name: Nick Haupka
    affiliation: State and University Library Göttingen
    affiliation_url: https://www.sub.uni-goettingen.de/
  - name: Anne Hobert
    affiliation: State and University Library Göttingen
    affiliation_url: https://www.sub.uni-goettingen.de/
    orcid_id: 0000-0003-2429-2995
description: Our investigation into OpenAlex uncovered more than 4 million records containing conflicting open access information. To illustrate this issue, we re-implemented Unpaywall's methodology using OpenAlex data. The comparative analysis revealed a shift, with over 1 million journal articles published in 2023, previously labeled as "closed" in OpenAlex, being re-classified as "gold", "hybrid", "green", or "bronze".
draft: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  comment = "#>",
  collapse = TRUE,
  warning = FALSE,
  message = FALSE,
  echo = FALSE,
  fig.width = 6,
  fig.asp = 0.618,
  out.width = "99%",
  fig.align = "center",
  dpi = 300
)
options(scipen = 999, digits = 4)
knitr::knit_hooks$set(
  inline = function(x) {
    if (is.numeric(x)) {
      return(prettyNum(x, big.mark = ","))
    } else{
      return(x)
    }
  }
)
library(tidyverse)
library(bigrquery)
library(DBI)
library(patchwork)
library(ggsankey)


## Database connection

library(DBI)
con <- dbConnect(
  bigrquery::bigquery(),
  project = "subugoe-collaborative",
  dataset = "openalex",
  billing = "subugoe-collaborative"
)
```

Over the last few months, we have switched our data basis for open access analytics from [Unpaywall](https://unpaywall.org/) to [OpenAlex](https://openalex.org/). Both open scholarly data services are developed by [OurResearch](https://ourresearch.org/) and have a similar metadata format for describing open access full-texts. However, OpenAlex offers monthly data dumps, which we've found particularly helpful as the free snapshot releases from Unpaywall, which we used before, appear to have ceased as of March 2022.

While transitioning from Unpaywall to OpenAlex, we noticed a considerable number of around 4 million OpenAlex records with contradictory open access information. This blog post aims to explore this issue. To better understand it, we've re-implemented Unpaywall's open access classification using OpenAlex data, and compared it against OpenAlex's existing classification.

## What is the issue?

OpenAlex provides various methods for identifying open access literature. Within the [work object](https://docs.openalex.org/api-entities/works/work-object), the [`open_access`](https://docs.openalex.org/api-entities/works/work-object#open_access) and [`best_oa_location`](https://docs.openalex.org/api-entities/works/work-object#best_oa_location) nodes contain information about the open access status at the article level. The [sources object](https://docs.openalex.org/api-entities/sources), on the other hand, gives information about the open access model of a journal.

The issue we've identified is that [filtering for open access works with `is_oa:true` returns around 4 million records with the `oa_status` marked as `closed`](http://api.openalex.org/works?filter=open_access.is_oa:true,open_access.oa_status:closed), a discrepancy we find inconsistent with OpenAlex’s own documentation. 
Accordingly, OpenAlex follows Unpaywall’s methodology tagging openly available works ([`is_oa`](https://docs.openalex.org/api-entities/works/work-object#is_oa)) and qualifying their open access status ([`oa_status`](https://docs.openalex.org/api-entities/works/work-object#oa_status)) using the following values:

-	gold: Published in an open access journal.
-	green: Toll-access on the publisher landing page, but there is a free copy in an OA repository.
-	hybrid: Free under an open license in a toll-access journal.
-	bronze: Free to read on the publisher landing page, but without any identifiable license.
-	closed: All other articles.

## Understanding the issue

To better understand this issue, we analysed the most recent OpenAlex snapshot from October 2023. After importing this data into our [BigQuery data warehouse](https://subugoe.github.io/scholcomm_analytics/data.html), we created a subset with the focus on journal articles published since 2013, excluding retractions and non-scholarly content published in journals.


```{sql create_dataset, connection = con, echo=TRUE, cache=TRUE}
CREATE OR REPLACE TABLE
  subugoe-collaborative.resources.oalex_cr_journal_articles_13_23 AS (
  SELECT
  doi,
  publication_year,
  open_access,
  best_oa_location,
  sources.is_oa AS journal_is_oa,
  sources.is_in_doaj AS journal_is_in_doaj,
  sources.host_organization_name AS publisher_name
FROM
  `subugoe-collaborative.openalex.works`
LEFT JOIN
  `subugoe-collaborative.openalex.sources` AS sources
ON
  primary_location.source.id = sources.id
WHERE
  type_crossref = "journal-article"
  AND is_paratext = FALSE
  AND is_retracted = FALSE
  AND publication_year BETWEEN 2013
  AND 2023 )
```


We then analysed the open access prevalence over the years, aggregating the record counts across both  `is_oa` and `oa_status`.

```{sql, label=oa_stats, connection = con, output.var="oa_stats", echo=TRUE}
SELECT
  COUNT(DISTINCT doi) AS articles,
  publication_year,
  open_access.is_oa,
  open_access.oa_status
FROM
  `subugoe-collaborative.resources.oalex_cr_journal_articles_13_23`
GROUP BY
  open_access.is_oa,
  open_access.oa_status,
  publication_year
ORDER BY
  publication_year DESC
```


The resulting figure shows the distribution of open access evidence in OpenAlex over years. All possible open access status values as known from Unpaywall were also represented in OpenAlex. The figure also presents the number of records tagged as open (blue colored bar chart stacks) and closed (colored bar chart stacks) access in the `is_oa` node. Notably,  the bulk of contradictory open access information could be found in records representing journal articles published in 2023, with `r oa_stats |> filter(publication_year == "2023", is_oa == TRUE, oa_status == "closed") |> pull(articles)` articles tagged as open access, but assigned the open access status "closed". 

```{r layout="l-body-outset"}
p_is_closed <- oa_stats |>
  mutate(oa_status = factor(oa_status, levels = c("gold", "hybrid", "green", "bronze", "closed"))) |>
  filter(oa_status == "closed") |>
  ggplot(aes(as.character(publication_year), articles, fill = is_oa)) +
  geom_bar(stat = "identity") +
  scale_fill_manual("is_oa", values = c(`FALSE` = "#b3b3b3a0", `TRUE` = "#56B4E9")) +
  scale_y_continuous(expand = expansion(mult = c(0, 0.05)), labels = scales::comma) +
  facet_wrap(~oa_status) +
  theme_minimal() +
  labs(y = "Articles", x = NULL)

p_not_closed <- oa_stats |>
  mutate(oa_status = factor(oa_status, levels = c("gold", "hybrid", "green", "bronze", "closed"))) |>
  filter(oa_status != "closed") |>
  ggplot(aes(as.character(publication_year), articles, fill = is_oa)) +
  geom_bar(stat = "identity") +
  scale_fill_manual("is_oa", values = c(`FALSE` = "#b3b3b3a0", `TRUE` = "#56B4E9")) +
  scale_y_continuous(expand = expansion(mult = c(0, 0.05)), labels = scales::comma) +
  facet_wrap(~oa_status) +
  theme_minimal() +
 # theme(axis.text.x = element_text(angle = 45, vjust = 0.5, hjust=1)) +
  labs(y = NULL, x = NULL)

design = "AB"
wrap_plots(
  A = p_is_closed,
  B = p_not_closed,
  design = design,
  widths = c(1, 2),
  heights = c(1, 1),
  guides = "collect"
) &
  theme(
    legend.position = "top",
    legend.justification = "right",
    legend.direction = "horizontal",
    panel.grid.minor = element_blank(),
                panel.border = element_rect(color = "grey50", fill = NA),
    text=element_text(size = 10),
    axis.text.x = element_text(angle = 45, hjust = 1),
    axis.title.x = element_text(margin = margin(t = 10)))
```


## Re-classification and analysis of changes

In order to address this inconsistency, we re-implemented [Unpaywall's open access classification methodology](https://support.unpaywall.org/support/solutions/articles/44001777288-what-do-the-types-of-oa-status-green-gold-hybrid-and-bronze-mean-). The following SQL code snippet shows how we approached the re-classification. 


```{sql, label=reclassify, connection=con, echo=TRUE}
CREATE OR REPLACE TABLE
  `subugoe-collaborative.resources.oalex_reclassify_oa` AS (
  SELECT
    DISTINCT doi,
    publication_year,
    open_access.is_oa,
    open_access.oa_status,
    CASE
      WHEN best_oa_location IS NULL THEN "closed"
      WHEN best_oa_location.source.type = "repository" THEN "green"
      WHEN (journal_is_in_doaj = TRUE OR journal_is_oa = TRUE) THEN "gold"
      WHEN (journal_is_in_doaj = FALSE
      AND journal_is_oa = FALSE )
    AND best_oa_location.license IS NOT NULL THEN "hybrid"
      WHEN (journal_is_in_doaj = FALSE AND journal_is_oa = FALSE ) AND best_oa_location.license IS NULL THEN "bronze"
    ELSE
    NULL
  END
    AS oa_new
  FROM
    `subugoe-collaborative.resources.oalex_cr_journal_articles_13_23` )
```

We used the best_oa_location node to determine the availability of an open access full-text. If this node was absent, we categorised the work as 'closed.' For open access works not exclusively provided by a repository, we used open access journal information from the source object to distinguish between gold, hybrid and bronze.

Following the re-classification, we calculated updated open access statistics.

```{sql, connection=con, output.var="oa_cat_changes", echo=TRUE}
SELECT
  COUNT(DISTINCT doi) AS n,
  oa_status,
  oa_new,
  publication_year
FROM
  `subugoe-collaborative.resources.oalex_reclassify_oa`
GROUP BY
  oa_status,
  oa_new,
  publication_year
```

The following figure compares OpenAlex open access classification (black-colored bars) with our approach (pink-colored bars). Notably, the re-classification resulted in many journal articles published in 2023 that were previously tagged as closed having one of the open access values such as gold, hybrid, green, or bronze.

```{r layout="l-body-outset"}
after <- oa_cat_changes |> 
  group_by(oa_status = oa_new, publication_year) |> 
  summarise(n = sum(n)) |>
  mutate(cat = "after")
before <- oa_cat_changes |> 
  group_by(oa_status= oa_status, publication_year) |> 
  summarise(n = sum(n)) |>
  mutate(cat = "before")
cat_df <- bind_rows(before, after) |>
  mutate(cat = factor(cat, levels = c("before", "after"))) |>
  mutate(oa_status = factor(oa_status, levels = c("gold", "hybrid", "green", "bronze", "closed")))


p_is_closed_new <- cat_df |>
  filter(oa_status == "closed") |>
  ggplot(aes(as.character(publication_year), n, fill = cat)) +
  geom_bar(stat = "identity", position = position_dodge2(preserve = "single")) +
  scale_fill_manual(NULL, values = c(`after` = "#fc5185", `before` = "#4d4d4d")) +
  scale_y_continuous(expand = expansion(mult = c(0, 0.05)), labels = scales::comma) +
  facet_wrap(~oa_status) +
  theme_minimal() +
  labs(y = "Articles", x = NULL)

p_is_open_new <- cat_df |>
  filter(oa_status != "closed") |>
  ggplot(aes(as.character(publication_year), n, fill = cat)) +
  geom_bar(stat = "identity", position = position_dodge2(preserve = "single")) +
  scale_fill_manual(NULL, values = c(`after` = "#fc5185", `before` = "#4d4d4d")) +
  scale_y_continuous(expand = expansion(mult = c(0, 0.05)), labels = scales::comma) +
  facet_wrap(~oa_status) +
  theme_minimal() +
  labs(y = NULL, x = NULL)

design = "AB"
wrap_plots(
  A = p_is_closed_new,
  B = p_is_open_new,
  design = design,
  widths = c(1, 2),
  heights = c(1, 1),
  guides = "collect"
) &
  theme(
    legend.position = "top",
    legend.justification = "right",
    legend.direction = "horizontal",
    panel.grid.minor = element_blank(),
                panel.border = element_rect(color = "grey50", fill = NA),
    text=element_text(size = 10),
    axis.text.x = element_text(angle = 45, hjust = 1),
    axis.title.x = element_text(margin = margin(t = 10)))
```

Overall, we re-classified a total of `r oa_cat_changes |> filter(!is.na(oa_new)) |> filter(oa_status != oa_new) |> pull(n) |> sum()` records representing journal articles published since 2013, with `r oa_cat_changes |> filter(publication_year == 2023, !is.na(oa_new)) |> filter(oa_status != oa_new) |> pull(n) |> sum()` of them being published in 2023. 
The following figure demonstrates changes in open access status after our re-classification for 2023.
The "gold" category gained `r  oa_cat_changes |> filter(publication_year == 2023, !is.na(oa_new)) |> filter(oa_status != oa_new, oa_new == "gold") |> pull(n) |> sum()` additional records in 2023, "hybrid" gained `r oa_cat_changes |> filter(publication_year == 2023, !is.na(oa_new)) |> filter(oa_status != oa_new, oa_new == "hybrid") |> pull(n) |> sum()`, "green" gained `r oa_cat_changes |> filter(publication_year == 2023, !is.na(oa_new)) |> filter(oa_status != oa_new, oa_new == "green") |> pull(n) |> sum()`, and `bronze` gained `r oa_cat_changes |> filter(publication_year == 2023, !is.na(oa_new)) |> filter(oa_status != oa_new, oa_new == "bronze") |> pull(n) |> sum()`. The figure also highlights that we not only re-assigned records previously assigned to the 'closed' category, but we could also observe changes between other categories.


```{r, cache=TRUE}
my_df <- oa_cat_changes |>
  filter(publication_year == 2023, !is.na(oa_new)) |>
  uncount(n) |>
  rename(before = oa_status, after = oa_new) |>
  make_long(before, after)
```

```{r}
my_df |>
  mutate(across(contains("node"), ~ factor(.x, levels = rev(
    c("gold", "hybrid", "green", "bronze", "closed")
  )))) |>
  ggplot(aes(
    x = x,
    next_x = next_x,
    node = node,
    next_node = next_node,
    fill = node,
    label = node
  )) +
  geom_sankey(flow.alpha = .6,
              node.color = "gray30") +
  scale_fill_manual("OA Status",
    values = c(
      `gold` = "#F5C710",
      `hybrid` = "#56B4E9",
      `green` = "#009E73",
      `bronze` = "#E69F00",
      `closed` = "#b3b3b3a0"
    ),
    guide = guide_legend(reverse = TRUE)
  ) +
  theme_sankey() +
  labs(x = NULL)
```

## Discussion and conclusion

Analysing and re-classifying open access data in OpenAlex has revealed inconsistencies in the actual implementation.

In response, we are sharing this detailed problem description in order to contribute to the ongoing improvement of OpenAlex, a scholarly data source that we enjoy working with on a daily basis. As a practical suggestion in the meantime, we recommend not to rely solely on the open access information provided. Instead, we suggest re-classifying open access status information.

## Funding {.appendix}

This work is funded by the Bundesministerium für Bildung und Forschung (BMBF) projects KBMINE (16WIK2101F) and KBOPENBIB (16WIK2301E). We acknowledge the support of the [German Competence Center for Bibliometrics](https://bibliometrie.info/).

