---
title: "Recent Changes in Document type classification in OpenAlex compared to Web of Science and Scopus"
description: In June 2024, we published a preprint on the classification of document types in OpenAlex and compared it with the scholarly databases Web of Science, Scopus, PubMed and Semantic Scholar. In this follow-up study, we want to investigate further developments in OpenAlex and compare the results with the proprietary databases Scopus and Web of Science. 
author:
  - name: Nick Haupka 
    affiliation: State and University Library Göttingen
    affiliation_url: https://www.sub.uni-goettingen.de/
    orcid_id: 0009-0002-6478-6789
  - name: Sophia Dörner
    affiliation: State and University Library Göttingen
    affiliation_url: https://www.sub.uni-goettingen.de/
    orcid_id: 0000-0001-8747-3422
  - name: Najko Jahn
    affiliation: State and University Library Göttingen
    affiliation_url: https://www.sub.uni-goettingen.de/
    orcid_id: 0000-0001-5105-1463
date: "`r Sys.Date()`"
output: distill::distill_article
bibliography: literature.bib
draft: TRUE
css: vis.css
---

```{r, echo = FALSE, message = FALSE, warning = FALSE}
knitr::opts_chunk$set(
  comment = "#>",
  collapse = TRUE,
  warning = FALSE,
  message = FALSE,
  echo = FALSE
)
options(scipen = 999, digits = 2)
knitr::knit_hooks$set(
  inline = function(x) {
    if (is.numeric(x)) {
      return(prettyNum(x, big.mark = ","))
    } else{
      return(x)
    }
  }
)
```

```{r, echo=FALSE,message = FALSE, warning = FALSE}
library(here)
library(tidyverse)
library(networkD3)
library(htmlwidgets)
library(reactable)
library(htmltools)
library(sankeyD3) # Not on CRAN, but here <https://github.com/fbreitwieser/sankeyD3>
```

```{r, echo=FALSE, message=FALSE, warning=FALSE}
oal2324_scp_t <- read_csv(here("_posts/openalex_document_types/data","oal2324_scp_t.csv"))
oal2324_wos_t <- read_csv(here("_posts/openalex_document_types/data","oal2324_wos_t.csv"))
scp_oal2324 <- read_csv(here("_posts/openalex_document_types/data","scp_oal2324.csv"))
wos_oal2324 <- read_csv(here("_posts/openalex_document_types/data","wos_oal2324.csv"))
```

In June 2024, we submitted an analysis of publication and document types in OpenAlex in comparison with the proprietary databases Web of Science and Scopus and the open data sources Semantic Scholar and PubMed [@haupka_2024].
We found substantial differences between these databases: While Web of Science and Scopus provided a comprehensive set of document types to describe works published in journals, OpenAlex supported only a comparably limited number of types.
Notably, OpenAlex lacked a distinction between research articles and reviews, which can be crucial when calculating citation indicators. 
In line with related studies [@alperin_analysis_2024], we also observed discrepancies in the number of publications when restricting to certain document types.

Meanwhile, in [late May]((https://groups.google.com/g/openalex-users/c/YujaIIjY02A)) and [late July 2024](https://groups.google.com/g/openalex-users/c/eXiWOlBXKC0), OpenAlex introduced extended approaches to obtain publication and document types.
Among the four new categories were preprints and reviews. Using [PubMed](https://groups.google.com/g/openalex-users/c/eXiWOlBXKC0), OpenAlex identified approximately 4 million journal articles as editorials, erratum, letters, preprints, reviews, or retractions. 

Of course, we wanted to know how these improvements affect our findings.
We therefore re-applied our approach to the recent changes.
Using works published in journals between 2012 and 2022, we demonstrate that OpenAlex's recent changes provide a more nuanced set of document types to refine scholarly works. 
However, the comparison with Web of Science and Scopus reveals that there remain considerable differences. 

## Data and Methods

Following our [preprint](https://arxiv.org/abs/2406.15154), we performed a pairwise comparison of journal publications indexed in OpenAlex with the Web of Science and Scopus published 2012 to 2022.
To investigate changes made in OpenAlex, we furthermore compared data from the OpenAlex July 2024 and August 2023 snapshots.
Scopus and Web of Science data were retrieved from the [German Competence Network of Bibliometrics](https://bibliometrie.info), using the April 2024 snapshots.
Web of Science data retrieval comprised the Core Collection. 
We matched items between the databases by DOI after normalisation to lowercase.
Overall, the intersection of OpenAlex and Scopus covered `r sum(oal2324_scp_t$n)` and the intersection of OpenAlex and Web of Science covered `r sum(oal2324_wos_t$n)` records.

Then, we categorised works based on their document type information into two categories: research discourse and editorial discourse. 
The research discourse category now also includes publications of type "preprint", which was added to OpenAlex in [May 2024](https://groups.google.com/g/openalex-users/c/YujaIIjY02A).
The mapping tables used for reclassifying the document types can be found in the appendix of @haupka_2024.

## Findings

Figure \@ref(fig:oalscp) illustrates OpenAlex document type changes in comparison with Scopus.
Before the introduction of the more nuanced set of document types, OpenAlex tagged
 `r sum(unique(oal2324_scp_t[oal2324_scp_t$oal23_type == "ARTICLE", ]$noal))` items (`r round(sum(unique(oal2324_scp_t[oal2324_scp_t$oal23_type == "ARTICLE", ]$noal)) / sum(unique(oal2324_scp_t$noal)) * 100, 2)`%) as articles, which reduced to `r sum(unique(oal2324_scp_t[oal2324_scp_t$oal24_type == "article", ]$noal))` (`r round(sum(unique(oal2324_scp_t[oal2324_scp_t$oal24_type == "article", ]$noal)) / sum(unique(oal2324_scp_t$noal)) * 100, 2)`%). 
Scopus tagged `r sum(unique(oal2324_scp_t[oal2324_scp_t$scp_type == "Article", ]$n24scp))` items (`r round(sum(oal2324_scp_t[oal2324_scp_t$scp_type == "Article", ]$n) / sum(unique(oal2324_scp_t$n24scp)) * 100, 2)`%) as article. 
OpenAlex assigned the type review to `r sum(unique(oal2324_scp_t[oal2324_scp_t$oal24_type == "review", ]$n24scp))` items (`r round(sum(unique(oal2324_scp_t[oal2324_scp_t$oal24_type == "review", ]$n24scp)) / sum(unique(oal2324_scp_t$n24scp)) * 100, 2)`%), whereas Scopus to `r sum(unique(oal2324_scp_t[oal2324_scp_t$scp_type == "Review", ]$n24scp))` items (`r round(sum(unique(oal2324_scp_t[oal2324_scp_t$scp_type == "Review", ]$n24scp)) / sum(unique(oal2324_scp_t$n24scp)) * 100, 2)`%). 


```{r oalscp, echo=FALSE,message=FALSE, warning=FALSE, fig.cap="Comparison of OpenAlex and Scopus for publication years 2012-2022"}
nodes <- oal2324_scp_t %>%
  group_by(oal23_type,oal24_type,scp_type) %>%
  select(oal23_type,oal24_type,scp_type) %>%
  unlist() %>%
  unique() %>%
  enframe("name","label") %>%
  mutate(name = as.character(name))

links <- bind_rows(
         oal2324_scp_t %>% 
         mutate(source=as.character(oal23_type), target=as.character(oal24_type), value=noal, .keep="none") %>%
         distinct(),
         oal2324_scp_t %>% 
         mutate(source=as.character(oal24_type), target=as.character(scp_type), value=n24scp, .keep="none") %>%
         distinct()) %>%
         mutate(source = match(source, nodes$label) -1,
                target = match(target, nodes$label) -1)

# Gruppen für jeden Knoten erstellen
nodes$group <- as.factor(ifelse(nodes$name %in% c(1,2,3,4,5,6,7), "oal23", 
                                ifelse(nodes$name %in% c(8,9,10,11,12,13,14,15,16), "oal24", "scp")))

# Farben für die source und target Knoten definieren
colours <- 'd3.scaleOrdinal() .domain(["oal23", "oal24","scp"]) .range(["#ffffff","#4d4d4d", "#d82222a0"])'

oal_scp_sn <- sankeyNetwork(Links = links, Nodes = nodes, Source = "source",
              Target = "target", Value = "value", NodeID = "label", fontSize = 12, nodePadding = 20, 
                    zoom = TRUE, showNodeValues = FALSE, numberFormat = ",d", fontFamily = "sans-serif", iterations=0,NodeGroup="group", colourScale = colours,dragX = TRUE,dragY = TRUE, xAxisDomain = c("OpenAlex (before)","OpenAlex (after)","Scopus"),xScalingFactor = .9, height=700)

oal_scp_sn <- onRender(
  oal_scp_sn,
  '
  function(el,x){
    var lowerCase = /^[a-z]/;
    var upperCase = /^[A-Z]/;
    var allCaps = /^[A-Z-]+$/;

    var labels = d3.select(el).selectAll(".node text");

    labels.each(function(d) {
      d.isAllCaps = allCaps.test(d.name);
      d.isLowerCase = lowerCase.test(d.name);
      d.isUpperCase = upperCase.test(d.name);

      let text = d.name.toLowerCase();
      let capitalized = text.replace(/(\\b|-)\\w/g, function(char) {
        return char.toUpperCase();
      });

      d3.select(this).text(capitalized + " (" + d3.format(",d")(d.value) + ")");
    });

    labels.attr("transform", function(d) {
        var box = this.getBBox();
        var adjust = box.width 
        if (d.isAllCaps) {
          return "translate(" + (-adjust - 40) + ",0)";
        }
        else if (d.isUpperCase) {
          return "translate(5,0)";
        }
        else {
          return "translate(5,0)";
        }
    });

    var zoomLayer = el.querySelector(".zoom-layer");
    if(zoomLayer){
      zoomLayer.setAttribute("transform", "translate(84,83) scale(0.75)");
    }

    labels.filter(function(d) { return d.isAllCaps; })
      .attr("x",x.options.nodeWidth - 10)
      .attr("text-anchor", "end")

    labels.filter(function(d) { return d.isLowerCase; })
      .attr("x", x.options.nodeWidth + 10)
      .attr("text-anchor", "start")
    
    labels.filter(function(d) { return d.isUpperCase; })
      .attr("x", x.options.nodeWidth + 10)
      .attr("text-anchor", "start")
    
    d3.select(el)
    .style("margin-left", "auto")
    .style("margin-right", "auto")
    .style("display", "block");
    
  }
  '
)
oal_scp_sn
```

Figure \@ref(fig:oalwos) illustrates the same for the comparison of OpenAlex with Web of Science. 
Here, OpenAlex tagged `r sum(unique(oal2324_wos_t[oal2324_wos_t$oal23_type == "ARTICLE", ]$noal))` items (`r round(sum(unique(oal2324_wos_t[oal2324_wos_t$oal23_type == "ARTICLE", ]$noal)) / sum(unique(oal2324_wos_t$noal)) * 100, 2)`%) as articles before the introduction of the more nuanced set of document types and `r sum(unique(oal2324_wos_t[oal2324_wos_t$oal24_type == "article", ]$noal))` (`r round(sum(unique(oal2324_wos_t[oal2324_wos_t$oal24_type == "article", ]$noal)) / sum(unique(oal2324_wos_t$noal)) * 100, 2)`%) after.
In Web of Science `r sum(oal2324_wos_t[oal2324_wos_t$wos_type == "Article", "n24wos"][!duplicated(oal2324_wos_t[oal2324_wos_t$wos_type == "Article", c("oal24_type", "wos_type", "n24wos")]),])` items (`r round(sum(oal2324_wos_t[oal2324_wos_t$wos_type == "Article", "n24wos"][!duplicated(oal2324_wos_t[oal2324_wos_t$wos_type == "Article", c("oal24_type", "wos_type", "n24wos")]),]) / sum(oal2324_wos_t$n24wos[!duplicated(oal2324_wos_t$n24wos)]) * 100, 2)`%) were tagged as articles. 
The document type review is assigned to `r sum(unique(oal2324_wos_t[oal2324_wos_t$oal24_type == "review", ]$n24wos))` items (`r round(sum(unique(oal2324_wos_t[oal2324_wos_t$oal24_type == "review", ]$n24wos)) / sum(unique(oal2324_wos_t$n24wos)) * 100, 2)`) by OpenAlex, whereas Web of Science tagged `r sum(unique(oal2324_wos_t[oal2324_wos_t$wos_type == "Review", ]$n24wos))` items (`r round(sum(unique(oal2324_wos_t[oal2324_wos_t$wos_type == "Review", ]$n24wos)) / sum(unique(oal2324_wos_t$n24wos)) * 100, 2)`%) as such. 

```{r oalwos, echo=FALSE,message=FALSE, warning=FALSE,fig.cap="Comparison of OpenAlex and Web of Science for publication years 2012-2022"}
nodes <- oal2324_wos_t %>%
  group_by(oal23_type,oal24_type,wos_type) %>%
  select(oal23_type,oal24_type,wos_type) %>%
  unlist() %>%
  unique() %>%
  enframe("name","label") %>%
  mutate(name = as.character(name))

links <- bind_rows(
         oal2324_wos_t %>% 
         mutate(source=as.character(oal23_type), target=as.character(oal24_type), value=noal, .keep="none") %>%
         distinct(),
         oal2324_wos_t %>% 
         mutate(source=as.character(oal24_type), target=as.character(wos_type), value=n24wos, .keep="none")  %>%
         distinct()) %>%
         mutate(source = match(source, nodes$label) -1,
                target = match(target, nodes$label) -1)

# Gruppen für jeden Knoten erstellen
nodes$group <- as.factor(ifelse(nodes$name %in% c(1,2,3,4,5,6,7), "oal23", 
                                ifelse(nodes$name %in% c(8,9,10,11,12,13,14,15,16,17), "oal24", "wos")))

# Farben für die source und target Knoten definieren
colours <- 'd3.scaleOrdinal() .domain(["oal23", "oal24","wos"]) .range(["#ffffff","#4d4d4d", "#9e0059a0"])'

oal_wos_sn <- sankeyNetwork(Links = links, Nodes = nodes, Source = "source",
              Target = "target", Value = "value", NodeID = "label", fontSize = 12, nodePadding = 20, 
                    zoom = TRUE, showNodeValues = FALSE, numberFormat = ",d", fontFamily = "sans-serif", iterations=0,NodeGroup="group", colourScale = colours,dragX = TRUE,dragY = TRUE, xAxisDomain = c("OpenAlex (before)","OpenAlex (after)","Web of Science"),xScalingFactor = .9, height=700)

oal_wos_sn <- onRender(
  oal_wos_sn,
  '
  function(el,x){
    var lowerCase = /^[a-z]/;
    var upperCase = /^[A-Z]/;
    var allCaps = /^[A-Z-]+$/;

    var labels = d3.select(el).selectAll(".node text");

    labels.each(function(d) {
      d.isAllCaps = allCaps.test(d.name);
      d.isLowerCase = lowerCase.test(d.name);
      d.isUpperCase = upperCase.test(d.name);

      let text = d.name.toLowerCase();
      let capitalized = text.replace(/(\\b|-)\\w/g, function(char) {
        return char.toUpperCase();
      });

      d3.select(this).text(capitalized + " (" + d3.format(",d")(d.value) + ")");
    });

    labels.attr("transform", function(d) {
        var box = this.getBBox();
        var adjust = box.width 
        if (d.isAllCaps) {
          return "translate(" + (-adjust - 40) + ",0)";
        }
        else if (d.isUpperCase) {
          return "translate(5,0)";
        }
        else {
          return "translate(5,0)";
        }
    });

    var zoomLayer = el.querySelector(".zoom-layer");
    if(zoomLayer){
      zoomLayer.setAttribute("transform", "translate(79,90) scale(0.75)");
    }

    labels.filter(function(d) { return d.isAllCaps; })
      .attr("x",x.options.nodeWidth - 10)
      .attr("text-anchor", "end")

    labels.filter(function(d) { return d.isLowerCase; })
      .attr("x", x.options.nodeWidth + 10)
      .attr("text-anchor", "start")
    
    labels.filter(function(d) { return d.isUpperCase; })
      .attr("x", x.options.nodeWidth + 10)
      .attr("text-anchor", "start")
    
    d3.select(el)
    .style("margin-left", "auto")
    .style("margin-right", "auto")
    .style("display", "block");
    
  }
  '
)
oal_wos_sn
```


Overall, Figures \@ref(fig:oalscp) and \@ref(fig:oalwos) demonstrate that even after the introduction of a more nuanced set of document types, OpenAlex still tags a higher proportion of items as articles than the commercial data sources.
The difference between the proportions of items tagged as articles is, however, slightly more pronounced in the comparison of OpenAlex with Web of Science.
Scopus tags a higher proportion of items as reviews and both Scopus and Web of Science still tag more items as editorial content than OpenAlex. 
In sum, `r sum(oal2324_scp_t[oal2324_scp_t$oal24_type == "article" & oal2324_scp_t$scp_type %in% c("Editorial", "Letter"), ]$n)` (Scopus) and `r sum(oal2324_wos_t[oal2324_wos_t$oal24_type == "article" & oal2324_wos_t$wos_type %in% c("Editorial Material", "Letter"), ]$n)` (Web of Science) items are tagged as editorial/editorial material or letters in Scopus and Web of Science, respectively, while tagged as articles in OpenAlex.

Here, the document types contained in each database were aggregated in two overarching categories: research discourse and editorial discourse. 
Both tables show that even after the introduction of a more nuanced set of document types in OpenAlex, the proportion of items labelled as editorial discourse is still about 3% lower compared to Scopus and Web of Science. 

<p/>
<p/>

```{r, echo=FALSE, message=FALSE, warning=FALSE}
react_bar_chart <-
  function(label,
           width = "100%",
           height = "14px",
           fill = "#00bfc4",
           background = NULL) {
    bar <-
      htmltools::div(style = list(
        background = fill,
        width = width,
        height = height
      ))
    chart <-
      htmltools::div(
        style = list(
          flexGrow = 1,
          marginLeft = "6px",
          background = background
        ),
        bar
      )
    htmltools::div(
      style = list(
        display = "flex",
        alignItems = "center",
        justifyContent = "flex-end"
      ),
      label,
      chart
    )
  }
```


```{r,layout='l-body-outset', echo=FALSE, message=FALSE, warning=FALSE}
scp_oal2324_reactable <- reactable(scp_oal2324,
  pagination = TRUE,
  highlight = TRUE,
  defaultColDef = colDef(vAlign = "center", headerClass = "header", headerVAlign = "center", align = "center"),
  compact = TRUE,
  columns = list(
     name = colDef(name="", align = "left",minWidth = 80),
     rd_total = colDef(name = "Total",
                       format = colFormat(separators = TRUE, locales = "en-GB"),
      width = 150,
      style = list(whiteSpace = "pre"),
      class = "number  border-left",
      cell = function(value, index) {
        width <- paste0(value * 100 / max(scp_oal2324$rd_total), "%")
        value <- format(value, big.mark = ",")
        value <- format(value, width = 10, justify = "right")
        bar <- div(
          class = "bar-chart",
          style = list(marginRight = "6px"),
          div(
            class = "bar",
            style = list(width = width, backgroundColor = "#c3c3c3")
          )
        )
        ### bring it all together, with the 'value' (number) preceding the bar itself
        div(class = "bar-cell", span(class = "number", value), bar)
      },
      align = "center"
    ),
    rd_pct = colDef(name = "Pct", format = colFormat(percent = TRUE, digits = 2),
                             cell = function(value) {
        value <- paste0(format(round(value * 100, 2), nsmall = 1), "%")
        value <- format(value, width = 5, justify = "right")
        react_bar_chart(value,
                        width = value,
                        fill = "#000000",
                        background = "#c3c3c3")
      },
      align = "center",
      style = list(whiteSpace = "pre"),
      class = "number",
      html = TRUE,
      minWidth = 100
    ),
     ed_total = colDef(name = "Total",
                       format = colFormat(separators = TRUE, locales = "en-GB"),
      width = 150,
      style = list(whiteSpace = "pre"),
      class = "number border-left",
      cell = function(value, index) {
        width <- paste0(value * 100 / max(scp_oal2324$ed_total), "%")
        value <- format(value, big.mark = ",")
        value <- format(value, width = 10, justify = "right")
        bar <- div(
          class = "bar-chart",
          style = list(marginRight = "6px"),
          div(
            class = "bar",
            style = list(width = width, backgroundColor = "#c3c3c3")
          )
        )
        ### bring it all together, with the 'value' (number) preceding the bar itself
        div(class = "bar-cell", span(class = "number", value), bar)
      },
      align = "center"
    ),
    ed_pct = colDef(name = "Pct", format = colFormat(percent = TRUE, digits = 2),
                             cell = function(value) {
        value <- paste0(format(round(value * 100, 2), nsmall = 1), "%")
        value <- format(value, width = 5, justify = "right")
        react_bar_chart(value,
                        width = value,
                        fill = "#000000",
                        background = "#c3c3c3")
      },
      align = "center",
      style = list(whiteSpace = "pre"),
      class = "number",
      html = TRUE,
      minWidth = 100
    ),
    na_total = colDef(name = "Total",
                       format = colFormat(separators = TRUE, locales = "en-GB"),
      width = 150,
      style = list(whiteSpace = "pre"),
      class = "number border-left",
      cell = function(value, index) {
        width <- paste0(value * 100 / max(scp_oal2324$na_total), "%")
        value <- format(value, big.mark = ",")
        value <- format(value, width = 10, justify = "right")
        bar <- div(
          class = "bar-chart",
          style = list(marginRight = "6px"),
          div(
            class = "bar",
            style = list(width = width, backgroundColor = "#c3c3c3")
          )
        )
        ### bring it all together, with the 'value' (number) preceding the bar itself
        div(class = "bar-cell", span(class = "number", value), bar)
      },
      align = "center"
    ),
    na_pct = colDef(name = "Pct", format = colFormat(percent = TRUE, digits = 3),
                             cell = function(value) {
        value <- paste0(format(round(value * 100, 3), nsmall = 1), "%")
        value <- format(value, width = 5, justify = "right")
        react_bar_chart(value,
                        width = value,
                        fill = "#000000",
                        background = "#c3c3c3")
      },
      align = "center",
      style = list(whiteSpace = "pre"),
      class = "number",
      html = TRUE,
      minWidth = 100
    )),
  # Create column groups
  columnGroups = list(
    colGroup(
      name = "Research Discourse",
      columns = c("rd_total", "rd_pct"),
      headerClass = "group-header"
    ),
    colGroup(
      name = "Editorial Discourse",
      columns = c("ed_total", "ed_pct"),
      headerClass = "group-header"
    ),
    colGroup(
      name = "Not Assigned",
      columns = c("na_total", "na_pct"),
      headerClass = "group-header"
    )
  ),
  searchable = FALSE,
  defaultPageSize = 10,
  showPageSizeOptions = TRUE
  )
scp_oal2324_reactable$sizingPolicy$padding <- 4
scp_oal2324_reactable
```

<p/>

```{r, layout='l-body-outset', echo=FALSE, message=FALSE, warning=FALSE}
wos_oal2324_reactable <- reactable(wos_oal2324,
  pagination = TRUE,
  highlight = TRUE,
  defaultColDef = colDef(vAlign = "center", headerClass = "header", headerVAlign = "center", align = "center"),
  compact = TRUE,
  columns = list(
     name = colDef(name="", align = "left",minWidth = 80),
     rd_total = colDef(name = "Total",
                       format = colFormat(separators = TRUE, locales = "en-GB"),
      width = 150,
      style = list(whiteSpace = "pre"),
      class = "number border-left",
      cell = function(value, index) {
        width <- paste0(value * 100 / max(wos_oal2324$rd_total), "%")
        value <- format(value, big.mark = ",")
        value <- format(value, width = 10, justify = "right")
        bar <- div(
          class = "bar-chart",
          style = list(marginRight = "6px"),
          div(
            class = "bar",
            style = list(width = width, backgroundColor = "#c3c3c3")
          )
        )
        ### bring it all together, with the 'value' (number) preceding the bar itself
        div(class = "bar-cell", span(class = "number", value), bar)
      },
      align = "center"
    ),
    rd_pct = colDef(name = "Pct", format = colFormat(percent = TRUE, digits = 2),
                             cell = function(value) {
        value <- paste0(format(round(value * 100, 2), nsmall = 1), "%")
        value <- format(value, width = 5, justify = "right")
        react_bar_chart(value,
                        width = value,
                        fill = "#000000",
                        background = "#c3c3c3")
      },
      align = "center",
      style = list(whiteSpace = "pre"),
      class = "number",
      html = TRUE,
      minWidth = 100
    ),
     ed_total = colDef(name = "Total",
                       format = colFormat(separators = TRUE, locales = "en-GB"),
      width = 150,
      style = list(whiteSpace = "pre"),
      class = "number border-left",
      cell = function(value, index) {
        width <- paste0(value * 100 / max(wos_oal2324$ed_total), "%")
        value <- format(value, big.mark = ",")
        value <- format(value, width = 10, justify = "right")
        bar <- div(
          class = "bar-chart",
          style = list(marginRight = "6px"),
          div(
            class = "bar",
            style = list(width = width, backgroundColor = "#c3c3c3")
          )
        )
        ### bring it all together, with the 'value' (number) preceding the bar itself
        div(class = "bar-cell", span(class = "number", value), bar)
      },
      align = "center"
    ),
    ed_pct = colDef(name = "Pct", format = colFormat(percent = TRUE, digits = 2),
                             cell = function(value) {
        value <- paste0(format(round(value * 100, 2), nsmall = 1), "%")
        value <- format(value, width = 5, justify = "right")
        react_bar_chart(value,
                        width = value,
                        fill = "#000000",
                        background = "#c3c3c3")
      },
      align = "center",
      style = list(whiteSpace = "pre"),
      class = "number",
      html = TRUE,
      minWidth = 100
    ),
    na_total = colDef(name = "Total",
                       format = colFormat(separators = TRUE, locales = "en-GB"),
      width = 150,
      style = list(whiteSpace = "pre"),
      class = "number border-left",
      cell = function(value, index) {
        width <- paste0(value * 100 / max(wos_oal2324$na_total), "%")
        value <- format(value, big.mark = ",")
        value <- format(value, width = 10, justify = "right")
        bar <- div(
          class = "bar-chart",
          style = list(marginRight = "6px"),
          div(
            class = "bar",
            style = list(width = width, backgroundColor = "#c3c3c3")
          )
        )
        ### bring it all together, with the 'value' (number) preceding the bar itself
        div(class = "bar-cell", span(class = "number", value), bar)
      },
      align = "center"
    ),
    na_pct = colDef(name = "Pct", format = colFormat(percent = TRUE, digits = 3),
                             cell = function(value) {
        value <- paste0(format(round(value * 100, 3), nsmall = 1), "%")
        value <- format(value, width = 5, justify = "right")
        react_bar_chart(value,
                        width = value,
                        fill = "#000000",
                        background = "#c3c3c3")
      },
      align = "center",
      style = list(whiteSpace = "pre"),
      class = "number",
      html = TRUE,
      minWidth = 100
    )),
  # Create column groups
  columnGroups = list(
    colGroup(
      name = "Research Discourse",
      columns = c("rd_total", "rd_pct"),
      headerClass = "group-header"
    ),
    colGroup(
      name = "Editorial Discourse",
      columns = c("ed_total", "ed_pct"),
      headerClass = "group-header"
    ),
    colGroup(
      name = "Not Assigned",
      columns = c("na_total", "na_pct"),
      headerClass = "group-header"
    )
  ),
  searchable = FALSE,
  defaultPageSize = 10,
  showPageSizeOptions = TRUE
  ) 
wos_oal2324_reactable$sizingPolicy$padding <- 4
wos_oal2324_reactable
```

## Discussion and Outlook
Our updated analysis demonstrated a noticable improvement of the classification of document types in OpenAlex when comparing it to Scopus and Web of Science. 
Compared to data from 2023, the discrepancy in the classification of items has decreased slightly. 
This indicates a convergence of the classification system in OpenAlex towards those from proprietary databases, with an enhanced coverage of reviews and editorial materials. 
In addition, the [rule-based string matching](https://github.com/ourresearch/openalex-guts/blob/main/detective/work_type_detective.py) for recognising paratexts introduced and revised by OpenAlex resulted in more texts being categorised as editorial material than before. 
However, the results also show that the curation of document types has not yet been finalised.

Conclusively, we would like to point out that there is no correct classification system per se. 
Rather different classification systems applied by the database operators can bring advantages and disadvantages. 
In Semantic Scholar and PubMed, for example, publications are labelled as clinical studies and case reports, which in Scopus, Web of Science and OpenAlex are predominantly assigned to the document type article. 
A differentiation of these publications has the potential to increase the quality of bibliometric surveys in the analysed databases. 
Also, the results from this analysis are only partially comparable with the results from our preprint, as in the preprint we worked with a more restrictive set that included publications from Semantic Scholar and PubMed.

## Funding {.appendix}

This work is funded by the Bundesministerium für Bildung und Forschung (BMBF) project KBOPENBIB (16WIK2301E). We acknowledge the support of the [German Competence Center for Bibliometrics](https://bibliometrie.info/).
